---
title: "Make Countries Move"
author: "Anton Bach Møller"
date: "05/1/2025"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore global development with R
In this exercise, you will load a filtered `gapminder` dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks within this script.** 

- **Submit to Github this script with solutions and its knitted version.**  
- **Submit to Brightspace a document with the (1) URL of this script and its knitted version in Github (2) answer of 250 words and a (3) visualisation of the Final Question**

## Get the necessary packages
First, start with installing and activating the relevant packages `tidyverse`, `gganimate`, and `gapminder` if you do not have them already. 
*Pay attention to what warning messages you get when installing `gganimate`, as your computer might need other packages than `gifski` and `av`*

```{r libraries, echo = FALSE}
# install.packages("gganimate")
# install.packages("gifski")
# install.packages("av")
# install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)
```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
head(gapminder)
```

The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() +
  ggtitle("Figure 01")
```

...

We see an interesting spread with an outlier to the right. Explore who it is so you can answer question 2 below!


Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() +
  ggtitle("Figure 02")
```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

### Questions for the static figures:

1. *Answer: why does it make sense to have a log10 scale (`scale_x_log10()`) on the x axis?* (hint: try to comment it out and observe the result, as in put , around the 
  scale_x_log10() + like this: ,  scale_x_log10() +,. It makes the graph very flat on the x axis)
  
  Since it takes into account the exponential growth of human development in the timeperiod we are working with. Thus making the graph more legible

2. *Answer: In Figure 1: Who is the outlier (the richest country in 1952) far right on the x axis?* 
The following code should show a table with the richest countries in descending order with the single most prosporus/the outlier being kuwait who is a whole order of magnitude richer in terms of gdp per capita.
```{ra}
# Filter the data for the year 1952
gapminder_1952 <- gapminder %>% filter(year == 1952)

# Sort the data by GDP per capita in descending order
gapminder_1952_sorted <- gapminder_1952 %>% arrange(desc(gdpPercap))

# View the sorted data
head(gapminder_1952_sorted)
```

3. *Fix Figures 1 and 2: Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation". You want to eliminate it.)*
```{rb}
# Load the scales package for formatting
library(scales)

# Plot for 1952 with scientific notation removed
ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() +
  scale_size_continuous(labels = label_number()) +  # Remove scientific notation for population
  ggtitle("Figure 01: 1952") +
  labs(x = "GDP per Capita (log scale)", y = "Life Expectancy", size = "Population")
```
```{rc}
# Plot for 1952 with continents colored and scientific notation removed
ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +  # Add transparency to make overlapping points visible
  scale_x_log10() +
  scale_size_continuous(labels = label_number()) +  # Remove scientific notation for population
  ggtitle("Figure 01: 1952") +
  labs(x = "GDP per Capita (log scale)", y = "Life Expectancy", size = "Population", color = "Continent")
```


4. *Answer: What are the five richest countries in the world in 2007?*
```{rd}
# Filter the data for the year 1952
gapminder_2007 <- gapminder %>% filter(year == 2007)

# Sort the data by GDP per capita in descending order
gapminder_2007_sorted <- gapminder_2007 %>% arrange(desc(gdpPercap))

# View the sorted data
head(gapminder_2007_sorted)
```



##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 starte}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10()  # convert x to log scale
anim
```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1f}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)

```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smooths the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2g}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim2
```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 

### Tasks for the animations:

5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)*

#to add a title to the previous animations I will use a function for each
 
```{r anim1h}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_size_continuous(labels = scales::label_number()) +  # Remove scientific notation
  labs(
    title = 'Year: {closest_state}',  # Dynamic title showing the year
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy",
    size = "Population",
    color = "Continent"
  ) +
  transition_states(year, transition_length = 1, state_length = 1)

# Render the animation
anim
```

```{r anim2i}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_size_continuous(labels = scales::label_number()) +  # Remove scientific notation
  labs(
    title = 'Year: {frame_time}',  # Dynamic title showing the year
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy",
    size = "Population",
    color = "Continent"
  ) +
  transition_time(year)

# Render the animation
anim2
```

#As can be seen in the two code chunks above, the "labs"/labels function change the labels of the graph.

6. *Can you made the axes' labels and units more readable? Consider expanding the abbreviated labels as well as the scientific notation in the legend and x axis to whole numbers. Also, differentiate the countries from different continents by color*

#I had kept the colouring from question 3 throughout, I am mostly just typing labels out

```{r anim1j}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::label_number()) +  # Remove scientific notation on x-axis
  scale_size_continuous(labels = scales::label_number()) +  # Remove scientific notation for population
  labs(
    title = 'Year: {closest_state}',  # Dynamic title showing the year
    x = "GDP per Capita (log scale)",  # Descriptive x-axis label
    y = "Life Expectancy (years)",     # Descriptive y-axis label
    size = "Population",               # Legend title for population
    color = "Continent"                # Legend title for continent
  ) +
  theme_minimal() +  # Use a clean theme for better readability
  transition_states(year, transition_length = 1, state_length = 1)

# Render the animation
anim
```

```{r anim2k}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::label_number()) +  # Remove scientific notation on x-axis
  scale_size_continuous(labels = scales::label_number()) +  # Remove scientific notation for population
  labs(
    title = 'Year: {frame_time}',  # Dynamic title showing the year
    x = "GDP per Capita (log scale)",  # Descriptive x-axis label
    y = "Life Expectancy (years)",     # Descriptive y-axis label
    size = "Population",               # Legend title for population
    color = "Continent"                # Legend title for continent
  ) +
  theme_minimal() +  # Use a clean theme for better readability
  transition_time(year)

# Render the animation
anim2
```

### Final Question
7. **Is the world a better place today than it was in the year you were born?** *Answer this question using the gapminder data.  Define better either as more prosperous, more free, more healthy, or suggest another measure that you can get from gapminder. Submit a 250 word answer with an illustration to Brightspace. Include a URL in your Brightspace submission that links to the coded solutions in Github. [Hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset or download more historical data at https://www.gapminder.org/data/ ]*

```{r anim1l}
# Load the additional life expectancy data
life_expectancy <- read.csv("C:/Users/Anton/OneDrive/Skrivebord/Rstudio 2025/data/life_expectancy.csv")

# View the first few rows of the data
head(life_expectancy)
```
```{r anim1m}
# View the first few rows of the data
head(life_expectancy)

# Check the column names
colnames(life_expectancy)

# Check the structure of the data
str(life_expectancy)
```
```{r anim1}
# View the first few rows of the data
head(life_expectancy)

# Check the column names
colnames(life_expectancy)

# Check the structure of the data
str(life_expectancy)
```

```{r anim1n}
# View the first few rows of the data
head(life_expectancy)

# Check the column names
colnames(life_expectancy)

# Check the structure of the data
str(life_expectancy)
```
#I encountered a problem, all years are numbered as x1800 etc so I have to remove the x somehow

#The following code reformats the data, only preserving two columns, year and lifexpectancy. So we lose the data of gdp etc since it is now in a "long" format instead of "wide".

```{r anim1o}
# Reshape the data from wide to long format
life_expectancy_long <- life_expectancy %>%
  pivot_longer(
    cols = starts_with("X"),  # Select all columns starting with "X" (years)
    names_to = "year",        # Create a new column called "year"
    values_to = "life_expectancy"  # Store the life expectancy values
  )

# View the reshaped data
head(life_expectancy_long)
```

#Now I remove the "X" that had been preventing me from using the original dataset. The "x" makes Rstudio recognise the column as charecters which prevents me from analysing.

```{r anim1p}
# Remove the "X" prefix and convert the year column to numeric
life_expectancy_long <- life_expectancy_long %>%
  mutate(year = as.numeric(sub("X", "", year)))

# View the cleaned data
head(life_expectancy_long)
```

#I am only interested in the data from 2000, the year I was born, to 2025 current year.

```{r anim1q}
# Filter the data for 2000 to 2025
life_expectancy_2000_2025 <- life_expectancy_long %>% filter(year >= 2000 & year <= 2025)

# View the filtered data
head(life_expectancy_2000_2025)
```

#Next we calculate the average lifespan of the global population from 2000 to 2025, as can be seen after running the code below, the global average lifespan is steadily increasing. There are some minor hicups though I wonder if a certain global pandemic had anything to do with the decline from 2019 to 2021.

```{r anim1r}
# Calculate average life expectancy for each year from 2000 to 2025
avg_life_exp <- life_expectancy_2000_2025 %>%
  group_by(year) %>%
  summarize(avg_life_exp = mean(life_expectancy, na.rm = TRUE))  # Remove NA values

# View the result
avg_life_exp
```

#Time to put an 
```{r anim1s}
# Line plot of average life expectancy from 2000 to 2025
ggplot(avg_life_exp, aes(x = year, y = avg_life_exp)) +
  geom_line(color = "blue", size = 1) +  # Add a blue line
  geom_point(color = "red", size = 3) +  # Add red points for each year
  labs(
    title = "Trend in Average Life Expectancy (2000-2025)",
    x = "Year",
    y = "Average Life Expectancy (years)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5))  # Add x-axis labels every 5 years
```

#Seems like we have just barely reached pre corona levels this year.

#below is the code I used to save the image to put into word

```{r anim1t}
# Create the plot
plot <- ggplot(avg_life_exp, aes(x = year, y = avg_life_exp)) +
  geom_line(color = "blue", size = 1) +  # Add a blue line
  geom_point(color = "red", size = 3) +  # Add red points for each year
  labs(
    title = "Trend in Average Life Expectancy (2000-2025)",
    x = "Year",
    y = "Average Life Expectancy (years)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5))  # Add x-axis labels every 5 years

# Print the plot to confirm it's created correctly
print(plot)
```

```{r anim1u}
# Save the plot as a PNG file
ggsave("life_expectancy_2000_2025.png", plot, width = 10, height = 6, dpi = 300)
```

