---
title: "Test on own data"
output: html_document
date: "2025-03-07"
---


Here is my submission, if you want to see it in R you can simply ctrl+alt+R and it will run everything, the error message below chunk that creates kings1-4 is simply because they convert different formats and mine only works with kings3

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE}
library(tidyverse)
read.csv("data/Spreadsheet of danish Kings.csv")
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator:the file is comma deliminated/seperated by commas

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/Spreadsheet of danish Kings.csv")

kings2 <- read_csv("data/Spreadsheet of danish Kings.csv")

kings3 <- read.csv2("data/Spreadsheet of danish Kings.csv")

kings4 <- read_csv2( "data/Spreadsheet of danish Kings.csv")
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
It is kings 2, still can't figure out how to know

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.
The class() function determines the class of the object it is describing, in this case
we get the answear of
[1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 
These are all abbreviations in order they mean:
Specified tibble dataframe
tibble data frame
and data frame isn't abbreviated. 
Thus we can see that they are tidyverse specific structures since they are tibbles
and they have meta data because the "spec_tbl_df" structure was needed which checks out

I have to not use kings2 due to using my own dataset that is danish instead of Adelas english excel sheet.


3. How many columns does the object have when created with these different functions?
11, found from running the ncol(kings) function
4. Show the dataset so that we can see how R interprets each column
I have used "glimps() to see how R interprets each column

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.
head(kings3)

kings <- kings3

class(kings)


ncol(kings)


glimpse(kings)


```



## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# Calculate the duration of reign
#I ended up just making a dataset without any dates attached, just year. I will have to learn lubridate if I want better for the exam.

kings3 <- kings3 %>%
  mutate(duration = End.reign.clean - Start.reign.clean)

# Inspect the updated dataset
glimpse(kings3)

# View the first few rows of the dataset
head(kings3)
# Calculate the average duration of reign
average_duration <- mean(kings3$duration)

# Print the average duration
print(paste("The average duration of reign is:", average_duration, "years"))
#So to answear the question, the average danish monarch reigns for 20.685185185185 

```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
# Filter kings with a longer-than-average duration of reign
long_reigning_kings <- kings3 %>%
  filter(duration > average_duration)

# Display the result
print(long_reigning_kings)

# Count the number of kings with a longer-than-average duration
long_reigning_count <- long_reigning_kings %>%
  count()

# Print the count
print(paste("Number of kings with a longer-than-average duration of reign:", long_reigning_count$n))

#running this chunk should give you the result of 27, which checks out since that is 50% of the total 54 danish kings
```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# Sort kings by duration in descending order and select the top 3
top_3_longest_ruling <- kings3 %>%
  arrange(desc(duration)) %>%
  slice(1:3)

# Display the result
print(top_3_longest_ruling)

# Print the total years ruled by the top 3
total_years <- sum(top_3_longest_ruling$duration)
print(paste("Total years ruled by the three longest-ruling monarchs:", total_years))
#In the object "top_3_longest_ruling" can we see Christian the 4th with 60, Margrethe the second with 52 and Christian the 9th with 43 years. I was told I could just use total clean years by Adela, so no days shall be counted.

```